Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-01-21 paper/current/main.tex
math section: sk

\documentclass[12pt]{article} %{{{

@@

% for quick author comments
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.8}
\def\del#1{ {\color{light-gray}{#1}} }
\def\yy#1{ {\color{red}\textbf{yy: #1}} }
\def\dk#1{ {\color{red}\textbf{dk: #1}} }
\def\js#1{ {\color{red}\textbf{js: #1}} }

@@

\subsection*{Mathematical connection between neural embedding model and gravity model}

Our results show that the neural embedding better explains a wide range of mobilities than conventional embedding methods, which prompts us to ask: why does the \textit{word2vec} work well for the mobility data?
We show that the \textit{word2vec} is tied to a gravity model, a widespread and long-standing model that well explains human mobility.

Let us approach the connection by initially focusing on the gravity model.
Suppose that flow of people between $N$ locations. The gravity model models the expected flow $\hat{T_{ij}}$ from location $i$ to $j$ by
\begin{align}
    \hat{T_{ij}} = C m_i m_j f(r_{ij}), \label{eq:gravity_model}
\end{align}
where $m_i$ is the population of locations of $i$, $r_{ij}$ is their distance, $f$ is a decreasing function of $r_{ij}$, and $C$ is a constant.
%In the following, we refer to a measure of dissimilarity as a distance for brevity.
%In gravity model, the flow is symmetric, \ie $\hat{T_{ij}}=\hat{T_{ji}}$, which ensures the stationarity in population $m_{i}$.

%Now, let us move our focus to the \textit{word2vec}.
The skip-gram negative sampling \textit{word2vec} takes a sequence of locations and models the co-occurrence of center location $i$ and context location $j$ by (See Methods)
\begin{align}
    P\left(j \given i \right):= \frac{p_0(j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}, \label{eq:prob_w2v_ng0}
\end{align}
where $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector", respectively,  $Z'_i=\sum_{x \in \mathcal{A}} p_0(x) \exp(\bm{u}_{x} \cdot \bm{v}_{i})$  is a normalization constant, and $\mathcal{A}$ is the set of all locations, and a noise distribution given by $p_{0}(j) = s^{\gamma} _j / \sum_{\ell'} s^{\gamma}_{\ell'}$, where $\gamma$ is a free parameter typically set to $3/4$ or $1$. Under restricting window size to be one, the expected flow between center location $i$ and context location $j$ is given by
\begin{align}
    \label{eq:flow_w2v}
    \hat{T_{ij}}=s_i P(j\given i) = \frac{s_i\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i},
\end{align}
where $s_i$ is the frequency of location $i$ summed for all trajectories in the dataset.

Parameter $\gamma=1$ is a special choice that ensures that the in-vector and out-vector are the same, \ie $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$, and substituting $\bm{u}_i = \bm{v}_j$ and $p_0(j) = s_j / \sum_{\ell'} s_{\ell'}$ into Eq~\eqref{eq:flow_w2v_ng0}, we have
\begin{align}
    P\left(j \given i \right) & := \frac{ p_0 (j) \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z'_i}                      \\
                              & =  \frac{s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z' _i\sum_{\ell'} s_{\ell'}},
\end{align}
which leads to the expected flow for the word2vec model, \ie
\begin{align}
    \hat{T_{ij}}=s_i P(j\given i) =  \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}}. \label{eq:flow_w2v_ng0}
\end{align}

In the \textit{word2vec}, the flow $\hat{T_{ij}}$ is symmetric by design (\ie $\hat{T_{ij}}=\hat{T_{ji}}$) because the model neglects whether the context $j$ appears later or earlier from center $i$.
Then, we have
\begin{align}
    T_{ij} = T_{ji}
     & \Rightarrow
    \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}} =  \frac{s_j s _i  \exp(\bm{v}_i \cdot \bm{v}_{j})}{  Z'_j\sum_{\ell'} s_{\ell'}} \\
     & \Rightarrow
    Z'_i = Z'_j, \label{eq:euqal_rel}
\end{align}
suggesting that the normalization constant $Z' _i$ is constant.
Combining Eqs.~\eqref{eq:flow_w2v_ng0} and \eqref{eq:euqal_rel}, we have
\begin{align}
    \hat{T_{ij}}=  C s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i}),
\end{align}
where $C$ is a positive constant.
Eq.~\eqref{eq:flow_w2v_ng0} reveals the connection between the \textit{word2vec} and the gravity model, with with $d_{ij} = - \bm{u}_{j} \cdot \bm{v}_{i}$ and $f(x) = \exp(-x)$.
In sum, the \textit{word2vec} is a gravity model, in which the population of location $i$ is its frequency in the trajectory data, and the distance between locations is measured by their dot similarities.

We also shows the reverse way of derivation (from gravity model to the \textit{word2vec}).
In the gravity model, the probability that a person moves from location $i$ to $j$ is given by
\begin{align}
    P\left(j \given i \right) = T_{ij} / m_i = C m_j f(d_{ij}). \label{eq:gravity_prob}
\end{align}

Because, the sum of probabilities should be one,
\begin{align}
     & \sum_{j} P(j \given i) = 1 \nonumber                                     \\
     & \Rightarrow C \sum_{k} m_k f(d_{ik}) = 1 \nonumber                       \\
     & \Rightarrow C  = 1 / \sum_{k} m_k f(d_{ik}), \label{eq:gravity_constant}
\end{align}
which ensures that $\sum_{j} m_j f(d_{ij})$ is constant.
Substituting \eqref{eq:gravity_constant}  into Eq.~\eqref{eq:gravity_prob} yields a model having the same functional form with the word2vec, \ie
\begin{align}
    P\left(j \given i \right) = \frac{m_i f(d_{ij})}{\sum_{k} m_k f(d_{ik})}. \label{eq:gravity_prob2}
\end{align}

>>>>

\documentclass[draft,12pt]{article} %{{{

@@

% for quick author comments
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.8}
\def\del#1{ {\color{light-gray}{#1}} }
\def\yy#1{ {\color{red}\textbf{yy: #1}} }
\def\dk#1{ {\color{red}\textbf{dk: #1}} }
\def\js#1{ {\color{red}\textbf{js: #1}} }
\def\sk#1{ {\color{red}\textbf{sk: #1}} }

@@

\subsection*{word2vec and gravity model}

Why does the embedding distance better predicts actual mobility flows than the geographic distance?
One plausible explanation is that an embedding is constructed based on mobility flows, making embedding distance useful for prediction.
However, there remains a question of why the embedding generated by the {\it word2vec} explains the flow exceptionally well.
We show that the {\it word2vec} can be translated into a gravity model, with mass and locations estimated from location trajectories.
It is this connection that gives rise to the {\it word2vec}'s exceptional ability to predict the flow.

Let us approach the question by initially focusing on how the {\it word2vec} generates the embedding.
The {\it word2vec} takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as an input.
A location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in its succeeding and proceeding $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
The {\it word2vec} learns an embedding by estimating a probability that location $i$ has context $j$ by
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where denominator $Z_i=\sum_{x \in \mathcal{A}} \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although the {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---which are referred to as in-vector and out-vector, respectively, many studies, as well as our experiments, adopt in-vector $\bm{v}_i$ as an embedding of location $i$.

Our results show that the {\it word2vec} best explains the mobility flow across different datasets when $w=1$, with the flow is predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T_{ij}}= P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(i)$ is the probability that location $i$ appears in the trajectory.
Apparently, there is no direct correspondence between the flow predicted by the {\it word2vec} and gravity model (Eq.~\eqref{eq:gravity_basic}).
The gap between the two models, however, can be filled by taking into account the effect of how the {\it word2vec} is trained.

A widespread training method for the {\it word2vec} is the {\it negative sampling} \cite{mikolov2013word2vec}.
The negative sampling finds the embedding vectors by solving a binary classification task, in which one is given a center-context location pair $(i, j)$ and classifies whether the pair comes from the data or an artificially generated ``fake'' data.
The fake pair $(i,j)$ is sampled based on the power $\gamma$ of the frequency of $j$, \ie $P^\gamma(j)$, where $\gamma$ is a parameter.
%typically sets to $\gamma=3/4$ or $\gamma = 1$.

The negative sampling does not fit the \textit{word2vec} directly to the data but indirectly through the synthetic classification task.
This has a side effect: the negative sampling has a bias and maximizes a different objective function~\autocite{Gutmann2010}.
When taking into account the bias, the {\it word2vec} with the negative sampling fits a probabiliy model given by
\begin{align}
	P\left(j \given i \right):= \frac{P^ \gamma(j) \exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng0}
\end{align}
where we have redefined the normalization constant, \ie $Z_i=\sum_{x \in \mathcal{A}} P^ \gamma (j) \exp(\bm{u}_{x} \cdot \bm{v}_{i})$. (See Skip-gram and word2vecc section for the full calculation).

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists  optimial in-vectors and out-vectors such that $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\bm{u}_i = \bm{v}_j$ into Eq.~\eqref{eq:prob_w2v_ng0}, the flow predicted by the {\it word2vec} is given by
\begin{align}
	\hat{T_{ij}}= P(i) P(j\given i) = \frac{P(i)P(j)\exp(\bm{v}_j \cdot \bm{v}_{i})}{Z_i}. \label{eq:flow_w2v_ng1}
\end{align}
The flow $\hat{T_{ij}}$ is symmetric (\ie $\hat{T_{ij}}=\hat{T_{ji}}$) because the model neglects whether the context $j$ appears later or earlier from center $i$.
In the righ-hand side of Eq.~\eqref{eq:flow_w2v_ng1}, the numerator remains the same even if we swap center and context locations $i$ and $j$.
Thus, to ensure the symmetry $T_{ij} = T_{ji}$, the denominator $Z_i$ should be a constant.

Taken altogether, the {\it word2vec} with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T_{ij}}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.
In other words, the \textit{word2vec} is a gravity model, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates the flow from the given mass and locations, the {\it word2vec} finds the mass and locations that best explain the given trajectories.
Therefore, the embedding distance is inherently tied to the flow, and hence, has greater predictive power than the geographic distance.
